---
title: "Hepatitis C Prediction"
author: "Trevor Okinda"
date: "2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Hepatitis C Prediction |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/fedesoriano/hepatitis-c-dataset\>*

### Reference:

*\<Lichtinghagen, R., Klawonn, F., & Hoffmann, G. (2015). Hepatitis C Prediction Dataset. UCI Machine Learning Repository. Retrieved from https://archive.ics.uci.edu/ml/datasets/HCV+data\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Exploratory Data Analysis
## Load dataset
```{r load dataset}
# Load dataset
hepatitis_data <- read.csv("HepatitisCdata.csv", colClasses = c(
  Category = "factor",
  Age = "numeric",
  Sex = "factor",
  ALB = "numeric",
  ALP = "numeric",
  ALT = "numeric",
  AST = "numeric",
  BIL = "numeric",
  CHE = "numeric",
  CHOL = "numeric",
  CREA = "numeric",
  GGT = "numeric",
  PROT = "numeric"
))

# Display the structure of the dataset
str(hepatitis_data)

# View the first few rows of the dataset
head(hepatitis_data)

# View the dataset in a separate viewer window
View(hepatitis_data)

```

## Measures of Frequency
```{r MOF}
# Frequency of each category
table(hepatitis_data$Category)

# Frequency of each sex
table(hepatitis_data$Sex)
```

## Measures of Central Tendency
```{r MOCT}
# Mean, median, and mode of ALB
mean_ALB <- mean(hepatitis_data$ALB)
median_ALB <- median(hepatitis_data$ALB)
mode_ALB <- names(sort(-table(hepatitis_data$ALB)))[1]

# Display results
cat("Mean ALB:", mean_ALB, "\n")
cat("Median ALB:", median_ALB, "\n")
cat("Mode ALB:", mode_ALB, "\n")

# Mean of other numerical variables
col_means <- colMeans(hepatitis_data[, sapply(hepatitis_data, is.numeric)])
col_means

```

## Measures of Distribution
```{r MOD}
# Summary statistics for numerical variables
summary(hepatitis_data[, sapply(hepatitis_data, is.numeric)])
```

## Measures of Relationship
```{r MOR}
# Correlation matrix
cor_matrix <- cor(hepatitis_data[, sapply(hepatitis_data, is.numeric)])
cor_matrix

```

## ANOVA
```{r ANOVA}
# Perform ANOVA for 'Category' against numerical variables
anova_results <- lapply(hepatitis_data[, sapply(hepatitis_data, is.numeric)], 
                        function(x) {
                          aov_out <- aov(x ~ Category, data = hepatitis_data)
                          return(summary(aov_out))
                        })

# Display ANOVA results
names(anova_results) <- colnames(hepatitis_data)[sapply(hepatitis_data, is.numeric)]
anova_results
```

## Plots
```{r Plots}
library(ggplot2)
library(GGally)

# Histogram for Age
ggplot(hepatitis_data, aes(x = Age)) +
  geom_histogram(binwidth = 5, fill = "skyblue", color = "black") +
  labs(title = "Histogram of Age", x = "Age", y = "Frequency")

# Boxplot for Age by Category
ggplot(hepatitis_data, aes(x = Category, y = Age, fill = Category)) +
  geom_boxplot() +
  labs(title = "Boxplot of Age by Category", x = "Category", y = "Age")

# Density plot for AST
ggplot(hepatitis_data, aes(x = AST)) +
  geom_density(fill = "skyblue", color = "black") +
  labs(title = "Density Plot of AST", x = "AST", y = "Density")

# Pairwise scatterplot matrix for numerical variables
pairs(hepatitis_data[, sapply(hepatitis_data, is.numeric)])

# Scatterplot of Age vs. AST colored by Category
ggplot(hepatitis_data, aes(x = Age, y = AST, color = Category)) +
  geom_point() +
  labs(title = "Scatterplot of Age vs. AST", x = "Age", y = "AST")


# Boxplot for Age
boxplot(hepatitis_data$Age)

# Histogram for Age
hist(hepatitis_data$Age)

# Scatterplot matrix for numerical variables
pairs(hepatitis_data[, sapply(hepatitis_data, is.numeric)])

```

# Preprocessing and Data Transformation
## Missing Values
```{r Missing Values}
# Check for missing values in the dataset
missing_values <- colSums(is.na(hepatitis_data))

# Print variables with missing values and their counts
print(missing_values[missing_values > 0])

# Display the total count of missing values in the dataset
total_missing <- sum(missing_values)
cat("Total missing values in the dataset:", total_missing, "\n")

```

## Dataset Cleaning
```{r Cleaning}
# Remove rows with missing values
hepatitis_data_clean <- na.omit(hepatitis_data)

# Display the dimensions of the cleaned dataset
cat("Dimensions of the cleaned dataset:", nrow(hepatitis_data_clean), "rows and", ncol(hepatitis_data_clean), "columns.\n")

# Check for missing values in the cleaned dataset
missing_values_clean <- colSums(is.na(hepatitis_data_clean))

# Print variables with missing values and their counts
print(missing_values_clean[missing_values_clean > 0])

# Display the total count of missing values in the cleaned dataset
total_missing_clean <- sum(missing_values_clean)
cat("Total missing values in the cleaned dataset:", total_missing_clean, "\n")
```

# Training Model
## Data Splitting
```{r Data Splitting}
library(caret)

# Set seed for reproducibility
set.seed(123)

# Create indices for train-test split (80% train, 20% test)
train_indices <- createDataPartition(hepatitis_data_clean$Category, p = 0.8, list = FALSE)

# Split the data into training and testing sets
train_data <- hepatitis_data_clean[train_indices, ]
test_data <- hepatitis_data_clean[-train_indices, ]

dim(train_data)
dim(test_data)

```

## Bootstrapping
```{r Bootstrapping}
library(boot)
# Define the function for bootstrapping
boot_function <- function(data, index) {
  return(mean(data[index, "AST"], na.rm = TRUE))
}

# Perform bootstrapping with 1000 resamples
set.seed(123)
bootstrap_results <- boot(data = train_data, statistic = boot_function, R = 1000)

# Summary of bootstrap results
summary(bootstrap_results)
```

## Model Training
```{r Model Training}
# Define the training control
train_control <- trainControl(method = "cv", number = 10)

# Train a model using cross-validation
# Example: Using Linear Discriminant Analysis (LDA)
library(MASS)  # For LDA
lda_model <- train(Category ~ ., data = train_data, method = "lda", trControl = train_control)

# Display cross-validation results
print(lda_model)

```

## Saving Model
```{r Saving Model}
# Load the saved LDA model
loaded_lda_model <- readRDS("./models/lda_model.rds")

# Prepare new data for prediction
new_data <- data.frame(
  Age = 32,
  Sex = "m",
  ALB = 38.5,
  ALP = 52.5,
  ALT = 7.7,
  AST = 22.1,
  BIL = 7.5,
  CHE = 6.93,
  CHOL = 3.23,
  CREA = 106,
  GGT = 12.1,
  PROT = 69
)

# Use the loaded model to make predictions for new data
predictions_loaded_model <- predict(loaded_lda_model, newdata = new_data)

# Print predictions
print(predictions_loaded_model)

```

